---
title: "Breaches During Covid"
author: "Matthew Webb"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About Project

This is a brief tutorial relating to ggplot and how to add annotations and text to a chart to help explain the visualization

First we need to load in the required packages


```{r}
library(httr)
library(tidyverse)
library(janitor)
library(readxl)
library(here)
library(gt)
library(ggplot2)
```

## Loading in the data

We need to source where our data is coming from, the below URL is from the NHS England Website and relates to the publicly available A&E Data

We also need to tell R what part of the workbook we wish to use for our analysis, and define the column types

```{r}
data_url <- 'https://view.officeapps.live.com/op/view.aspx?src=https%3A%2F%2Fwww.england.nhs.uk%2Fstatistics%2Fwp-content%2Fuploads%2Fsites%2F2%2F2023%2F10%2FMonthly-AE-Time-Series-September-2023.xls&wdOrigin=BROWSELINK'

download.file(url = data_url, destfile = here('data', 'Monthly-AE-Time-Series-September-2023.xlsx'), mode = 'wb')

Monthly_Data <- read_excel(here("Monthly-AE-Time-Series-September-2023.xls"), sheet = 1, range = "B14:N172",
  col_names = TRUE,
  col_types = c("date","numeric","numeric","numeric",
              "numeric","numeric","numeric", "numeric",
              "numeric","numeric","numeric", "numeric",
              "numeric"), skip = 13) |>
  clean_names()
```

## Formatting the columns

Next we need to format our columns and select the ones we wish to use

When you import data into R the date variables are defined as POSIX date, so we can format them into standard R date format

```{r formatted variables}
Monthly_Data_Formatted <- Monthly_Data |>
  # Formatting variables to integers and renaming to make them more manageable
  mutate(
    `Total Attendances` = as.integer(total_attendances),
    `Total Emergency Admissions` = as.integer(total_emergency_admissions),
    `4 Hour Admissions` = as.integer(number_of_patients_spending_4_hours_from_decision_to_admit_to_admission),
    `12 Hour Admissions` = as.integer(number_of_patients_spending_12_hours_from_decision_to_admit_to_admission),
    `Period Formatted` = as.Date(period),
    `Year` = year(`Period Formatted`)) |>
  # Keeping the formatted columns
  select(`Period Formatted`,
         `Year`,
         `Total Attendances`, 
         `Total Emergency Admissions`, 
         `4 Hour Admissions`, 
         `12 Hour Admissions`) |>
  view()
```

## Overview Table

It's useful to create an initial table to get an idea of the overall trends
We can group by the Year variable to get aggregate counts

```{r summary table}
Yearly_Table <- Monthly_Data_Formatted |>
  group_by(`Year`) |>
  summarise(
    `Total Emergency Admissions` = sum(`Total Emergency Admissions`),
    `4 Hour Admissions` = sum(`4 Hour Admissions`),
    `12 Hour Admissions` = sum(`12 Hour Admissions`)
  ) |>
  gt() |>
  tab_header(title = md("**Total A&E Breaches in England 2010-2023**"),
             subtitle = md("Both 4-Hour and 12-Hour Breaches")) |>
  tab_source_note(source_note = "Source: NHS England A&E Monthly Statistics") |>
  fmt_number(columns = c(`Year`, `Total Emergency Admissions`,`4 Hour Admissions`, `12 Hour Admissions`),
             sep_mark = "",
             decimals = 0)

Yearly_Table
```

## A simple plot with title labels

This is a simple plot showing total emergency admissions

```{r simple plot}
Simple_Plot <- ggplot(Monthly_Data_Formatted, aes(x = `Period Formatted`, y = `Total Emergency Admissions`)) +
  geom_line() +
  labs(title = "Total A&E Admissions in England",
       subtitle = "For 2010-2023",
       x = "Date",
       y = "Total A&E Admissions")
Simple_Plot

```

## Applying a scale to the x axis

Currently the x axis is only showing 3 years, we can change this by adding a scale

```{r simple plot with scale}
Simple_Plot_Dates <- ggplot(Monthly_Data_Formatted, aes(x = `Period Formatted`, y = `Total Emergency Admissions`)) +
  geom_line() +
  labs(title = "Total A&E Admissions in England",
       subtitle = "For 2010-2023",
       x = "Date",
       y = "Total A&E Admissions") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")

Simple_Plot_Dates
```

## Creating initial breaches plot

Now we can swap out the admissions variable for the breaches and add some colour to distinguish them

```{r simple breaches plot}
Simple_Breaches_Plot <- ggplot(Monthly_Data_Formatted, aes(x = `Period Formatted`, y = `4 Hour Admissions`)) +
  geom_line(colour = "red", size = 1) +
  geom_line(aes(x = `Period Formatted`, y = `12 Hour Admissions`), colour = "purple", size = 1) +
  labs(title = "Total A&E Breaches in England",
       subtitle = "For 2010-2023",
       x = "Date",
       y = "Total A&E Breaches") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")

Simple_Breaches_Plot
```
## Adding annotations

We can now add some annotations to help visualize this plot better

We can highlight the COVID period with a rectangle zone, so we need to know a few details about the pandemic:

COVID-19 was officially confirmed to be in the UK in January 2020

It spread throughout the UK and in July 2021 the delta variant began spreading 

So we can define the COVID period as being roughly from January 2020 to December 2021

```{r annotate function}
Annotated_Breaches_Plot <- ggplot(Monthly_Data_Formatted, aes(x = `Period Formatted`, y = `4 Hour Admissions`)) +
  geom_line(colour = "red", size = 1) +
  geom_line(aes(x = `Period Formatted`, y = `12 Hour Admissions`), colour = "purple", size = 1) +
  labs(title = "Total A&E Breaches",
       subtitle = "For 2010-2023",
       x = "Date",
       y = "Total A&E Breaches") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")+
  annotate('rect', xmin = as.Date("2020-01-01"), xmax = as.Date("2021-12-31"), ymin = 0, ymax = 130000,
           alpha = 0.1, fill = "blue", col = "black") +
  annotate('text', x = as.Date("2018-08-01"), y = 130000, label = "COVID Period")

Annotated_Breaches_Plot
```
## Adding an arrow 

Lets add an arrow to the label to make it clearer

```{r adding arrow}
Arrow_Breaches_Plot <- ggplot(Monthly_Data_Formatted, aes(x = `Period Formatted`, y = `4 Hour Admissions`)) +
  geom_line(colour = "red", size = 1) +
  geom_line(aes(x = `Period Formatted`, y = `12 Hour Admissions`), colour = "purple", size = 1) +
  labs(title = "Total A&E Breaches",
       subtitle = "For 2010-2023",
       x = "Date",
       y = "Total A&E Breaches") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")+
  annotate('rect', xmin = as.Date("2020-01-01"), xmax = as.Date("2021-12-31"), ymin = 0, ymax = 130000,
           alpha = 0.1, fill = "blue", col = "black") +
  annotate('text', x = as.Date("2018-08-01"), y = 130000, label = "COVID Period") + 
   annotate('curve', x = as.Date("2018-10-01"), xend = as.Date("2019-12-01"),
            y = 125000, yend = 110000, linewidth = 0.5, curvature = 0.5,
             arrow = arrow(length = unit(0.4,'cm')))

Arrow_Breaches_Plot
```

## Adding events to the COVID Period

There are some events that will help illustrate the COVID Period, they come from the government website:

https://www.gov.uk/government/publications/covid-19-and-occupational-impacts/covid-19-and-occupational-impacts#:~:text=Pandemic%20in%20the%20UK.,of%20SARS%2DCoV%2D2%20.

- National lockdowns were late March 2020 - June 2020 and January 2021 – July 2021
- First lockdown March 2020
- Second lockdown November 2020
- Third lockdown January 2021
- Most restrictions removed July 2021

```{r adding lines}
Line_Breaches_Plot <- ggplot(Monthly_Data_Formatted, aes(x = `Period Formatted`, y = `4 Hour Admissions`)) +
  geom_line(colour = "red", size = 1) +
  geom_line(aes(x = `Period Formatted`, y = `12 Hour Admissions`), colour = "purple", size = 1) +
  labs(title = "Total A&E Breaches",
       subtitle = "For 2010-2023",
       x = "Date",
       y = "Total A&E Breaches") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")+
  annotate('rect', xmin = as.Date("2020-01-01"), xmax = as.Date("2021-12-31"), ymin = 0, ymax = 130000,
           alpha = 0.1, fill = "blue", col = "black") +
  annotate('text', x = as.Date("2018-08-01"), y = 130000, label = "COVID Period") + 
   annotate('curve', x = as.Date("2018-10-01"), xend = as.Date("2019-12-01"),
            y = 125000, yend = 110000, linewidth = 0.5, curvature = 0.5,
             arrow = arrow(length = unit(0.4,'cm'))) +
  # Below is the code for adding the vertical lines
  geom_segment(aes(x = as.Date("2020-03-01"), y = 0, yend = 150000,
                     xend = as.Date("2020-03-01"), colour = "segment"),
                     linewidth = 1, colour="green") +
  geom_segment(aes(x = as.Date("2020-11-01"), y = 0, yend = 150000,
                     xend = as.Date("2020-11-01"), colour = "segment"),
                     linewidth = 1, colour="green") +
  geom_segment(aes(x = as.Date("2021-01-01"), y = 0, yend = 150000,
                     xend = as.Date("2021-01-01"), colour = "segment"),
                     linewidth = 1, colour="green") +
  annotate('text',x = as.Date("2018-05-01"),y = 160000,label = "First lockdown") +
    annotate('text',x = as.Date("2020-03-01"),y = 180000,label = "Second lockdown") +
    annotate('text',x = as.Date("2021-09-01"),y = 165000,label = "Third lockdown") +
  # Adding arrows
   annotate('curve',x = as.Date("2019-01-01"),xend = as.Date("2020-03-01"), y = 155000, yend = 140000 ,
    linewidth = 1,curvature = 0.6,arrow = arrow(length = unit(0.2,'cm'))) +
     annotate('curve',x = as.Date("2020-02-01"),xend = as.Date("2020-11-01"), y = 175000, yend = 150000,
    linewidth = 1,curvature = 0.6,arrow = arrow(length = unit(0.2,'cm'))) +
   annotate('curve',x = as.Date("2021-10-01"),xend = as.Date("2021-01-01"), y = 155000 , yend = 140000,
    linewidth = 1,curvature = 0.6,arrow = arrow(length = unit(0.2,'cm'))) 

Line_Breaches_Plot
```